---
title: "CodingNotes_LinearModels"
author: "Brynleigh Payne"
date: "2025-04-03"
output: 
  pdf_document:
  md_document:
    variant: gfm
  word_document:
  html_document:
    toc: true
    toc_float: true
---

Load in data
```{r}
data("mtcars")
```

Load packages
```{r}
library(ggplot2)
```

## Continuous X and Y
Plot data
```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_smooth(method = lm, se = FALSE, color = "grey") + # "lm" = linear model
  geom_point(aes(color = wt)) +
  xlab("Weight") + 
  ylab("Miles per gallon") +
  scale_colour_gradient(low = "forestgreen", high = "black") +
  theme_classic()
```
Run a linear model
```{r}
lm(mpg~wt, data = mtcars)
```
To get a pvalue from linear model use "summary" function
```{r}
summary(lm(mpg~wt, data = mtcars))
```
Run ANOVA (multiple categorical variables) on linear model
```{r}
anova(lm(mpg~wt, data = mtcars))
```
Run correlation analysis 
```{r}
cor.test(mtcars$wt, mtcars$mpg)
```
All tests gave the same pvalue because they are all basically running a linear regression! Wow!

## Assumptions
- y is continuous
- error is normally distributed 
- relationship is linear 
- homoskedasticity 
- sigma is consistent 
- independent samples

Visualize data to assess for assumptions broken
```{r}
model <- lm(mpg~wt, data = mtcars)

ggplot(model, aes(y = .resid, x = .fitted)) +
  geom_point() +
  geom_hline(yintercept = 0)

```

## Categorical Variables

Load packages
```{r}
library(tidyverse)
```

Read in data
```{r}
bull.rich <- read.csv("Bull_richness.csv")
```

Filter our dataset to include one treatment and growth stage for demonstration of a t-test
```{r}
bull.rich %>%
  filter(GrowthStage == "V8" & Treatment == "Conv.") %>%
  ggplot(aes(x = Fungicide, y = richness)) + 
  geom_boxplot()
```

Run a t-test
```{r}
bull.rich.sub <- bull.rich %>%
  filter(GrowthStage == "V8" & Treatment == "Conv.")

t.test(richness~Fungicide, data = bull.rich.sub)
```

We can plug this into our linear model and get the same results. Assuming equal variance in groups and performing a two-sample t-test is the same result as a linear model and ANOVA on the linear model. 
```{r}
t.test(richness~Fungicide, data = bull.rich.sub, var.equal = TRUE)
summary(lm(richness~Fungicide, data = bull.rich.sub))
anova(lm(richness~Fungicide, data = bull.rich.sub))
```
## ANOVA

Load packages
```{r}
#install.packages("emmeans")
library(emmeans)
library(multcomp)
#install.packages("multcompView")
library(multcompView)
```

Filter our dataset to have only richness in different crop growth stages in the control samples in soybean in conventional management
```{r}
bull.rich.sub2 <- bull.rich %>%
  filter(Fungicide == "C" & Treatment == "Conv." & Crop == "Corn")
```

Visualize data
```{r}
bull.rich.sub2$GrowthStage <- factor(bull.rich.sub2$GrowthStage, levels = c("V6", "V8", "V15"))

ggplot(bull.rich.sub2, aes(x = GrowthStage, y = richness)) +
  geom_boxplot()
```

Linear model 
```{r}
lm.growth <- lm(richness ~ GrowthStage, data = bull.rich.sub2)
summary(lm.growth)
```

ANOVA
```{r}
anova(lm.growth)
summary(aov(richness ~ GrowthStage, data = bull.rich.sub2))
```

We can do post-hoc tests to find out which groups are different from each other. The simplest way to think of this is individual t-tests across groups.
```{r}
lsmeans <- emmeans(lm.growth, ~GrowthStage) # estimate lsmeans of variety within siteXyear
Results_lsmeans <- cld(lsmeans, alpha = 0.05, reversed = TRUE, details = TRUE) # contrast with Tukey ajustment by default. 
Results_lsmeans

```

## Interaction Terms

Filter our dataset to include fungicide term
```{r}
bull.rich.sub3 <- bull.rich %>%
  filter(Treatment == "Conv." & Crop == "Corn")

bull.rich.sub3$GrowthStage <- factor(bull.rich.sub3$GrowthStage, levels = c("V6", "V8", "V15"))
```

Set up our linear model with fungicide interaction factor
```{r}
# write it like this with the :
lm.inter <- lm(richness ~ GrowthStage + Fungicide + GrowthStage:Fungicide, data = bull.rich.sub3)

# or like this with the *
lm(richness ~ GrowthStage*Fungicide, data = bull.rich.sub3)
```

ANOVA table
```{r}
summary(lm.inter) # significant terms
anova(lm.inter) # The interaction term is signifant. 
```

Plot the effect of fungicide over the levels of growthstage
```{r}
bull.rich.sub3 %>%
  ggplot(aes(x = GrowthStage, y = richness, fill = Fungicide)) +
  geom_boxplot()
```

Post-hoc test within the levels of growthstage
```{r}
lsmeans <- emmeans(lm.inter, ~Fungicide|GrowthStage) # estimate lsmeans of variety within siteXyear
Results_lsmeans <- cld(lsmeans, alpha = 0.05, reversed = TRUE, details = TRUE) # contrast with Tukey ajustment
Results_lsmeans
```

## Mixed Effects Model
Mixed effects models have both fixed effects and mixed effects. 

Our interaction model from above
```{r}
lme0 <- lm(richness ~ GrowthStage*Fungicide, data = bull.rich.sub3)
```

Load packages
```{r}
library(lme4)
```

Generalized Linear Model
```{r}
lme1 <- lmer(richness ~ GrowthStage*Fungicide + (1|Rep), data = bull.rich.sub3)
summary(lme1)
```

Look at what it does to our Betas in our original model
```{r}
summary(lme0)
```
Look at what it does to our Betas in our generalized linear model
```{r}
summary(lme1)
```

[Link to my github](https://github.com/bep0022/PLPA5820)
